How predictable is extinction? Forecasting species survival at million-year timescales

### 2019-07-09 doc/article.tex
seth discussion comments

\begin{abstract}
  A tenet of conservation palaeobiology is that knowledge of past extinction patterns can help us to better predict future extinctions. Although the future is unobservable, we can test the strength of this proposition by asking how well models conditioned on past observations would have predicted subsequent extinction events at different points in the geological past. To answer this question, we analyze the well-sampled fossil record of Cenozoic planktonic microfossil taxa (foramanifera, radiolarians, diatoms, and calcareous nanoplankton). We examine how extinction probability varies over time as a function of species age, time of observation, current geographic range, change in geographic range, climate state, and change in climate state. Our models have a 70-80\% probability of correctly forecast the rank order of extinction risk for a random out-of-sample species pair, implying that determinants of extinction have varied only modestly through time. We find that models which include either historical covariates or account for variation in covariate effects over time yield equivalent forecasts, but a model including both is overfit and yields biased forecasts. An important caveat is that human impacts may substantially disrupt range-risk dynamics so that the future will be less predictable than it has been in the past.

@@

Here we take this approach, using as a model system the Cenozoic record of skeletonized marine planktonic microorganisms (Foraminifera, Radiolaria, Diatoms, and Coccolithophores). This record has several key strengths for our purposes: planktonic microorganisms are widespread and abundant in pelagic habitats, have high preservation potential, and because of their utility for biostratigraphic, paleoclimatic, and oceanographic study they have been the focus of an extensive international coring and study effort \citep{Lazarus1994,SpencerCervato1999}. A compliation of these data is readily available through the Neptune database, an online repository of species occurrences obtained through the Deep Sea Drilling Program and the Ocean Drilling Project \citep{Lazarus1994,SpencerCervato1999}. This database provides abundant samples in space and time, a high degree of temporal resolution for the entirety of the Cenozoic, and has an taxonomic synonymization framework for dealing with 50+ years of taxonomic opinion \citep{Lazarus1994} -- as close to ideal data for this analysis as possible. Analyzing patterns of extinction and global occurrence at fine temporal scales means we can better elucidate how well we can predict species extinction at human-relevant scales.

@@

We analyzed microfossil occurrence information from the Neptune Database \url{http://www.nsb-mfn-berlin.de/nannotax} \citep{Lazarus1994,SpencerCervato1999}. This occurrence-based dataset includes calcareous nannoplankton, diatoms, planktonic foraminifera, and radiolarians. Occurrences were filtered to include only those species with first occurrences no earlier than 63 Mya (millions of years ago). This filtering criterion excludes taxa that survived the K/Pg extinction or arose during this recovery interval, and ensures that our occurrence histories fully overlap with the temperature time-series used as a potential extinction risk predictor (see below).

@@

\begin{table}[ht]
 \caption{Models and their definitions}
 \begin{threeparttable}
  {
   \def\arraystretch{1.5}
   \begin{tabular}{ l p{3cm} l l }
    Code & Description & Covariates & R Formula Syntax\tnote{a}\phantom{\textsuperscript{a}} \\
    \hline
    C & Constant effects, no historical cov. & \parbox[t]{0.25\textwidth}{Geographic range,\\temperature} & \parbox[t]{0.33\textwidth}{event\tnote{b}\phantom{\textsuperscript{b}} $\sim$ range\tnote{c}\phantom{\textsuperscript{c}} + temp\tnote{d}\phantom{\textsuperscript{d}} +\\(1 $|$ time\tnote{e}\phantom{\textsuperscript{e}}/phylum\tnote{f}\phantom{\textsuperscript{f}}) +\\(1 $|$ age\tnote{g}\phantom{\textsuperscript{g}}/phylum)} \\
    V & Varying effects, no historical cov. & \parbox[t]{0.25\textwidth}{Geographic range,\\temperature} & \parbox[t]{0.33\textwidth}{event $\sim$ range + temp +\\(1 + range + temp $|$ time/phylum) + (1 $|$ age/phylum)} \\ 
    CP & Constant effects, historical cov. & \parbox[t]{0.25\textwidth}{Geographic range,\\change in geographic range, temperature,\\previous temperature} & \parbox[t]{0.33\textwidth}{event $\sim$ + range\_diff1\tnote{g}\phantom{\textsuperscript{h}} + range\_diff2\tnote{h}\phantom{\textsuperscript{h}} + range\_diff3\tnote{h}\phantom{\textsuperscript{h}} + \\temp + temp\_lag\tnote{i}\phantom{\textsuperscript{i}} +\\(1 $|$ time/phylum) +\\(1 $|$ age/phylum)} \\
    VP & Varying effects, historical cov. & \parbox[t]{0.25\textwidth}{Geographic range,\\change in geographic range, temperature,\\previous temperature} & \parbox[t]{0.33\textwidth}{event $\sim$ range + range\_diff1 + range\_diff2 + range\_diff3 + \\temp + temp\_lag +\\(1 + range + range\_diff1 + range\_diff2 + range\_diff3 +\\temp + temp\_lag $|$ time/phylum) +\\(1 $|$ age/phylum)} \\
    \hline
   \end{tabular}
  }
  \begin{tablenotes}
  \item[a] See Supplemental Equation S2 for full statistical model definition.
  \item[b] Species observation where 1 if time of last observation, otherwise 0.
  \item[c] Species geographic range in log km\(^2\). Mean centered, scaled to sd = 1.
  \item[d] Global temperature in degrees C. Mean centered, scaled to sd = 1.
  \item[e] Time of observation.
  \item[f] Taxonomic group of species (i.e. Foraminifera, Diatoms, Radiolarians, Calcareous nannoplankton).
  \item[g] Age at observation.
  \item[h] Change in geographic range since last observation (number indicates how lags).
  \item[i] Temperature at previous observation.
  \end{tablenotes}
 \end{threeparttable}
 \label{tab:model_def}
\end{table}

@@

We are interested in our models' performance in two distinct contexts: in-sample performance, and out-of-sample predictive performance (i.e. forecasting).

@@

Cross-validation is a procedure for estimating a models expected out-of-sample error. Information criteria such as AIC or WAIC are approximations of out-of-sample predictive error as estimated by cross-validation \citep{ESL,Gelman2013d}. Cross-validation implicitly takes into account model complexity because when a model is overfit to its data, out-of-sample predictions will be biased and inaccurate \citep{ESL}. A high degree of similarity between out-of-sample and in-sample estimates indicates that the model is not overfit to the data (though it is not necessarily an adequate descriptor of the data). Cross-validation is preferable to simple metrics such as AIC because instead of a single value it produces, an entire posterior distribution of estimates.

@@

% ROC model comparison
\begin{figure}[ht]
 \begin{subfigure}[ht]{0.45\textwidth}
  \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/auc_hist_full}
  \caption{In-sample}
  \label{fig:auc_hist}
 \end{subfigure}
 \begin{subfigure}[ht]{0.45\textwidth}
  \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/fold_auc_full}
  \caption{Out-of-sample}
  \label{fig:fold_auc}
 \end{subfigure}
 \caption{Comparisons of measures of model performance for both in-sample (\ref{fig:auc_hist}) and out-of-sample (\ref{fig:fold_auc}) cross-validation. The area under the receiver operating charactered curve (AUC) was calculated for each model. these estimates are calculated from the models posterior predictive distribution (\ref{fig:auc_hist}) or from predictions made to new data (\ref{fig:fold_auc}), respectively. Marked below the posterior distributions are the median AUC and 50\% and 80\% posterior intervals for all observations in our dataset. Models with higher AUC values indicate better performance over models with lower AUC values. AUC is bounded between 0.5 and 1. See Table \ref{tab:model_def} for an explanation of the four models (C, V, CP, VP).}
 \label{fig:auc_compare}
\end{figure}

@@

However, the actual difference in forecast AUC result between the VP model and the other three models is extremely small (0.01 AUC unit), and all of the in-sample AUC estimates from our models are concentrated between an AUC value of 0.775 and 0.795 (Fig. \ref{fig:auc_hist}). This result indicates that the practical difference in performanc between these models might be so small that there is no practical benefit that the VP model over the other three. Ultimately, determining which of these models produces the best forecasts of future extinctions requires comparing these in-sample results to our out-of-sample results (see below).

The in-sample forecasts from our four models over time are broadly similar between taxonomic groups (Fig. \ref{fig:auc_taxon_time}). Our in-sample forecasts for Diatoms are the weakest of the taxonomic groups as all four models have an equal number of intervals with no predictive power (AUC approximately 0.5). In contrast, our best in-sample forecast results are for Radiolarians where for any of our models there is at most 1 interval with almost no predictive power. Ultimately, our in-sample forecasts over time by taxonimc group are broadly consistent between our four models.

@@

Expected out-of-sample forecasting performance was estimated using five-fold cross-validation for time series \citep{Arlot2010,Bergmeir2016}. Our out-of-sample forecast AUC estimates demonstrate a broader range of results, with AUC estimates ranging between approximately 0.7 and 0.85 (Fig. \ref{fig:auc_hist}, \ref{fig:fold_auc}). While our VP model was the model with greatest in-sample forecasting performance (Fig. \ref{fig:auc_hist}), this model has the greatest decrease in out-of-sample forecasting performance (Fig. \ref{fig:fold_auc}). In comparison, the other three models demontrstrate a much smaller difference between in-sample forecast and out-of-sample forecast AUC values which indicates that our VP model is overfit to our data, and that one of the simpler model would be preferable for predicting future extinctions. This result means that our model which includes the historical covariates (e.g. change in geographic range) and allowing these effects to vary over time produces biased extinciton forecasts. Interestingly, including either historical covariates like change in geographic range or allowing the effects of geographic range and other covariate effects to vary over time have approximately equal effect on forecasting future extinction events (Fig. \ref{fig:auc_compare}).

In the analysis of the in-sample forecast performance of the four models, we noted that there were time intervals where our predictions were no better than random (Fig. \ref{fig:auc_taxon_time}). This occurrence is generally much rarer for the posterior distribution of AUC from the out-of-sample forecasts. The major exception to this pattern are our estimates for the diatoms, which have at least one time interval for all four models in which the median AUC of the out-of-sample forecasts were no better random. The only other group for which median posterior predictive estimate of out-of-sample AUC reaches 0.5 is calcareous nannoplankton, and then only with the V model.
\begin{figure}[ht]
 \centering
 \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/fold_auc_taxon_time_full}
 \caption{Comparison of our models ability to forecast future extinction events as measured by out-of-sample AUC values over time as aggregated by taxonomic group for each of the four models. The AUC of the individual My intervals within each fold is plotted to highlight the heterogeneity in performance within and between folds. This presentation decomposes each of the 12-million year folds by each of the taxonomic groups into the predictions made for each of the million-year intervals. The black line corresponds to the median AUC estimate, with the envelopes corresponding to multiple credible intervals as indicated in the legend. The grey intervals mark the geologic ages of the Cenozoic. See Table \ref{tab:model_def} for a description of each of the four models (C, V, CP, VP).}
 \label{fig:fold_auc_taxon_time}
\end{figure}

We compared the difference in our AUC estimates from the out-of-sample forecasts to the AUC estimates from our in-sample forecasts by subtracting the in-sample AUC estimates from the out-of-sample AUC estimates (Fig. \ref{fig:oos_ins_diff}). A difference in AUC close to 0 indicates complete congruence between the in-sample and out-of-sample forecasts. A positive difference indicates that our out-of-sample forecasts are actually higher performing than our in-sample forecasts, while negative difference indicates poorer out-of-sample performance than in-sample forecast. Divergences between our out-of-sample and in-sample forecasts are rare and tend to not form multimillion year patters, consistent with the broad visual congruence between the in-sample and out-of-sample forecast performance (Fig. \ref{fig:auc_taxon_time}, \ref{fig:fold_auc_taxon_time}). The only major multimillion year pattern indicating significantly poorer out-of-sample forecast performance than in-sample forecast performance is for Radiolaria based on the VP model concentrated around 30 Mya (Fig. \ref{fig:oos_ins_diff}).

\begin{figure}[ht]
 \centering
 \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/auc_diff}
 \caption{Comparing our models' ability to forecast out-of-sample extinction compared to our models' in-sample forecasts. Congruence between these measures indicates that our models are not necessarily overfit to the data. This value is calculated as the values presented in Figure \ref{fig:fold_auc_taxon_time} minus those values presented in Figure \ref{fig:auc_taxon_time}. A differences close to 0 indicate complete congruence between in-sample and out-of-sample forecasts, while a positive difference indicates that our out-of-sample forecasts are actually higher performing than our in-sample forecasts, and a negative difference indicates poorer out-of-sample performance than in-sample forecast. See Table \ref{tab:model_def} for a description of each of the four models (C, V, CP, VP).}
 \label{fig:oos_ins_diff}
\end{figure}

@@

Three of the four models we evaluated are practically identical in their ability to make in-sample and out-of-sample forecasts. Although the in-sample AUC estimates differ between models, all of these estimates are in a narrow range of possible AUC values (Fig. \ref{fig:auc_hist}). The model with the best in-sample forecasting results includes the historical covariates and allows all covariate effects to vary over time. However, the out-of-sample forecasts from this model are biased forecasts, indicating that it is overfit to our data \ref{fig:fold_auc}). The model that includes historical covariates such as geographic range trajectory yields out-of-sample forecasts with nearly identical results to the model that allows covariate effects to vary over time but does not include historical covariates.

We note that both including historical covariates such as change in geographic range and allowing covariate effects to vary over time are different ways of encoding information from the past. Including historical covariates in a model but not allowing covariate effects to vary over time encodes the past explicitly but assumes that covariates effects are constant over time. Allowing covariate effects to vary over time, on the other hand, does not explicitly encode ``the past'' into our model but instead models how covariate effects vary over time which allows the past to be implicitly included in our model. By modeling this variation, forecasts made for future extinction events can allow for a wide range of potential effects of our covariates on predicting extinction. Comparing our out-of-sample forecast results indicates that these approaches yield approximately equal forecasting performance (Fig. \ref{fig:fold_auc}). Our results supplement those of \citet{Kiessling2016} which examined using differences in geographic range over time to predict extinction risk.

An extremely important caveat, of course, is that human impacts may substantially alter present and future extinction risk dynamics relative to the average Cenozoic state, so that the future may become less predicatable than it has been in the past \citep{Harnik2012a,Finnegan2015}. Our model with historical covariates assumes that extinction selectivity with respect to these covariates is constant through time, but given growing evidence that human impacts substantially alter extinction risk dynamics \citep{Harnik2012a,Finnegan2015,Payne2007}, this assumption may not be valid and may limit or bias our ability to predict extinction in truly novel environmental regimes. Thus, it might be preferable to model the variation in extinction risk and selectivity over time instead of relying solely on measures from a taxon's past such as change in geographic range over time. For this reason, while our CP and V models yield similar out-of-sample forecasts, we believe the V model offers more practical benefits for predicting extinction risk in future, anthropogenically impacted environments.

The relative quality and consistency between in-sample and out-of-sample forecasting performance for three of the four models is encouraging given that these estimates are based on very limited biological and environmental information about the studied taxa. Even our most complex models only account for a few simple aspects of geographic range, prior history, and phylogenetic affinity. The principal reason we were not able to include more biological information in the models used here is because we lack additional life history or ecological information for many of the marine micro- and nannoplankton included in this study. Foraminifera are an exception to this problem as aspects of life history, ecology, and physiology are known for many foram species \citep{Ezard2011}. However, comparable information does not exist for all foram species, nor does this type of data exist for the other three taxonomic groups studied here. Future analyses including this type of information and focused more narrowly on the foraminifera may be informative.

In summary, our results suggest that models trained on prior extinction/survival patterns do modestly well at predicting relative extinction probability of randomly selected species pairs based on a small number of simple taxonomic, geographic, and historical predictors. Although a model that includes historical covariates such as change in geographic range and change in climate between observations while also allowing covariate effects to vary over time performs best at in-sample prediction, this model is overfit to our data and produces less accurate out-of-sample forecasts than our three less complex models. The remaining three models yield nearly equivalent out-of-sample forecasts, suggesting that including historical information via either explicit modeling of historical covariate effects or modeling how covariate effects have changed over time does not diminish and may ultimately improve our ability to forecast future extinctions. The results of this simple exercise suggest that conservation decisions could indeed be bolstered by including fossil data.

>>>>

\begin{abstract}
  A tenet of conservation palaeobiology is that knowledge of past extinction patterns can help us to better predict future extinctions. Although the future is unobservable, we can test the strength of this proposition by asking how well models conditioned on past observations would have predicted subsequent extinction events at different points in the geological past. To answer this question, we analyze the well-sampled fossil record of Cenozoic planktonic microfossil taxa (Foramanifera, Radiolaria, diatoms, and calcareous nanoplankton). We examine how extinction probability varies over time as a function of species age, time of observation, current geographic range, change in geographic range, climate state, and change in climate state. Our models have a 70-80\% probability of correctly forecast the rank order of extinction risk for a random out-of-sample species pair, implying that determinants of extinction have varied only modestly through time. We find that models which include either historical covariates or account for variation in covariate effects over time yield equivalent forecasts, but a model including both is overfit and yields biased forecasts. An important caveat is that human impacts may substantially disrupt range-risk dynamics so that the future will be less predictable than it has been in the past.

@@

Here we take this approach, using as a model system the Cenozoic record of skeletonized marine planktonic microorganisms (Foraminifera, Radiolaria, diatoms, and calcareous nannoplankton including coccolithophores). This record has several key strengths for our purposes: planktonic microorganisms are widespread and abundant in pelagic habitats, have high preservation potential, and because of their utility for biostratigraphic, paleoclimatic, and oceanographic study they have been the focus of an extensive international coring and study effort \citep{Lazarus1994,SpencerCervato1999}. A compliation of these data is readily available through the Neptune database, an online repository of species occurrences obtained through the Deep Sea Drilling Program and the Ocean Drilling Project \citep{Lazarus1994,SpencerCervato1999}. This database provides abundant samples in space and time, a high degree of temporal resolution for the entirety of the Cenozoic, and has an taxonomic synonymization framework for dealing with 50+ years of taxonomic opinion \citep{Lazarus1994} -- as close to ideal data for this analysis as possible. Analyzing patterns of extinction and global occurrence at fine temporal scales means we can better elucidate how well we can predict species extinction at human-relevant scales.

@@

We analyzed microfossil occurrence information from the Neptune Database \url{http://www.nsb-mfn-berlin.de/nannotax} \citep{Lazarus1994,SpencerCervato1999}. This occurrence-based dataset includes calcareous nannoplankton, diatoms, planktonic Foraminifera, and radiolarians. Occurrences were filtered to include only those species with first occurrences no earlier than 63 Mya (millions of years ago). This filtering criterion excludes taxa that survived the K/Pg extinction or arose during this recovery interval, and ensures that our occurrence histories fully overlap with the temperature time-series used as a potential extinction risk predictor (see below).

@@

\begin{table}[ht]
 \caption{Models and their definitions}
 \begin{threeparttable}
  {
   \def\arraystretch{1.5}
   \begin{tabular}{ l p{3cm} l l }
    Code & Description & Covariates & R Formula Syntax\tnote{a}\phantom{\textsuperscript{a}} \\
    \hline
    C & Constant effects, no historical cov. & \parbox[t]{0.25\textwidth}{Geographic range,\\temperature} & \parbox[t]{0.33\textwidth}{event\tnote{b}\phantom{\textsuperscript{b}} $\sim$ range\tnote{c}\phantom{\textsuperscript{c}} + temp\tnote{d}\phantom{\textsuperscript{d}} +\\(1 $|$ time\tnote{e}\phantom{\textsuperscript{e}}/phylum\tnote{f}\phantom{\textsuperscript{f}}) +\\(1 $|$ age\tnote{g}\phantom{\textsuperscript{g}}/phylum)} \\
    V & Varying effects, no historical cov. & \parbox[t]{0.25\textwidth}{Geographic range,\\temperature} & \parbox[t]{0.33\textwidth}{event $\sim$ range + temp +\\(1 + range + temp $|$ time/phylum) + (1 $|$ age/phylum)} \\ 
    CP & Constant effects, historical cov. & \parbox[t]{0.25\textwidth}{Geographic range,\\change in geographic range, temperature,\\previous temperature} & \parbox[t]{0.33\textwidth}{event $\sim$ + range\_diff1\tnote{g}\phantom{\textsuperscript{h}} + range\_diff2\tnote{h}\phantom{\textsuperscript{h}} + range\_diff3\tnote{h}\phantom{\textsuperscript{h}} + \\temp + temp\_lag\tnote{i}\phantom{\textsuperscript{i}} +\\(1 $|$ time/phylum) +\\(1 $|$ age/phylum)} \\
    VP & Varying effects, historical cov. & \parbox[t]{0.25\textwidth}{Geographic range,\\change in geographic range, temperature,\\previous temperature} & \parbox[t]{0.33\textwidth}{event $\sim$ range + range\_diff1 + range\_diff2 + range\_diff3 + \\temp + temp\_lag +\\(1 + range + range\_diff1 + range\_diff2 + range\_diff3 +\\temp + temp\_lag $|$ time/phylum) +\\(1 $|$ age/phylum)} \\
    \hline
   \end{tabular}
  }
  \begin{tablenotes}
  \item[a] See Supplemental Equation S2 for full statistical model definition.
  \item[b] Species observation where 1 if time of last observation, otherwise 0.
  \item[c] Species geographic range in log km\(^2\). Mean centered, scaled to sd = 1.
  \item[d] Global temperature in degrees C. Mean centered, scaled to sd = 1.
  \item[e] Time of observation.
  \item[f] Taxonomic group of species (i.e. Foraminifera, diatoms, Radiolaria, calcareous nannoplankton).
  \item[g] Age at observation.
  \item[h] Change in geographic range since last observation (number indicates how lags).
  \item[i] Temperature at previous observation.
  \end{tablenotes}
 \end{threeparttable}
 \label{tab:model_def}
\end{table}

@@

We are interested in model performance (i.e. forecasting) in two distinct contexts: in-sample performance, and out-of-sample predictive performance.

@@

Cross-validation is a procedure for estimating a model's expected out-of-sample error. Information criteria such as AIC or WAIC are approximations of out-of-sample predictive error as estimated by cross-validation \citep{ESL,Gelman2013d}. Cross-validation implicitly takes into account model complexity because when a model is overfit to its data, out-of-sample predictions will be biased and inaccurate \citep{ESL}. A high degree of similarity between out-of-sample and in-sample estimates indicates that the model is not overfit to the data (though it is not necessarily an adequate descriptor of the data). Cross-validation is preferable to simple metrics such as AIC because instead of a single value it produces, an entire posterior distribution of estimates.

@@

% ROC model comparison
\begin{figure}[ht]
 \begin{subfigure}[ht]{0.45\textwidth}
  \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/auc_hist_full}
  \caption{In-sample}
  \label{fig:auc_hist}
 \end{subfigure}
 \begin{subfigure}[ht]{0.45\textwidth}
  \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/fold_auc_full}
  \caption{Out-of-sample}
  \label{fig:fold_auc}
 \end{subfigure}
 \caption{Comparisons of measures of model performance for both in-sample (\ref{fig:auc_hist}) and out-of-sample (\ref{fig:fold_auc}) cross-validation. The area under the receiver operating charactered curve (AUC) was calculated for each model. these estimates are calculated from the models posterior predictive distribution (\ref{fig:auc_hist}) or from predictions made to new data (\ref{fig:fold_auc}), respectively. Marked below the posterior distributions are the median AUC and 50\% and 80\% posterior intervals for all observations in the dataset. Models with higher AUC values indicate better performance over models with lower AUC values. AUC is bounded between 0.5 and 1. See Table \ref{tab:model_def} for an explanation of the four models (C, V, CP, VP).}
 \label{fig:auc_compare}
\end{figure}

@@

However, the actual difference in forecast AUC result between the VP model and the other three models is extremely small (0.01 AUC unit), and all of the in-sample AUC estimates from our models are concentrated between an AUC value of 0.775 and 0.795 (Fig. \ref{fig:auc_hist}). This result indicates that the practical difference in performance between these models might be so small that there is no practical benefit that the VP model over the other three. Ultimately, determining which of these models produces the best forecasts of future extinctions requires comparing these in-sample results to our out-of-sample results (see below).

In-sample forecasts from the four models over time are broadly similar between taxonomic groups (Fig. \ref{fig:auc_taxon_time}). In-sample forecasts for diatoms are the weakest of the four taxonomic groups as all four models have several intervals with no predictive power (AUC not significantly greater than 0.5). The best in-sample forecast results are for radiolarians, for which all models have at most 1 interval with little predictive power. In-sample forecasts over time by taxonomic group are broadly consistent between the four models.

@@

We estimated expected out-of-sample forecasting performance using five-fold cross-validation for time series \citep{Arlot2010,Bergmeir2016}. Out-of-sample forecast AUC estimates exhibit a broader range than in-sample estimates, with AUC ranging between approximately 0.7 and 0.85 (Fig. \ref{fig:auc_hist}, \ref{fig:fold_auc}). While the VP model has the best in-sample forecasting performance (Fig. \ref{fig:auc_hist}) this model performs poorly at out-of-sample forecasting compared to the other models (Fig. \ref{fig:fold_auc}). The poor out-of-sample performance suggests that this most complex model is overfit and that one of the simpler models would be preferable for predicting future extinctions. Thus the models that include both historical covariates (e.g. change in geographic range) and time-varying effects produce biased extinction forecasts. Interestingly, models that include either historical covariates but assume constant effects (CP) or do not include historical covariates but include time-varying effects (V) perform similarly when forecasting future extinction events (Fig. \ref{fig:auc_compare}).

As noted above there were some time intervals in which in-sample forecasts were no better than random (Fig. \ref{fig:auc_taxon_time}). Such intervals are generally much rarer for out-of-sample forecasts. The major exception to this pattern are diatoms, which have at least one time interval for all four models in which the median AUC of the out-of-sample forecasts were no better random. The only other group for which median posterior predictive estimate of out-of-sample AUC reaches 0.5 is calcareous nannoplankton, and then only with the V model.
\begin{figure}[ht]
 \centering
 \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/fold_auc_taxon_time_full}
 \caption{Comparison of the four models' ability to forecast future extinction events as measured by out-of-sample AUC values over time as aggregated by taxonomic group for each of the four models. The AUC of the individual My intervals within each fold is plotted to highlight the heterogeneity in performance within and between folds. This presentation decomposes each of the 12-million year folds by each of the taxonomic groups into the predictions made for each of the million-year intervals. The black line corresponds to the median AUC estimate, with the envelopes corresponding to multiple credible intervals as indicated in the legend. The grey intervals mark the geologic ages of the Cenozoic. See Table \ref{tab:model_def} for a description of each of the four models (C, V, CP, VP).}
 \label{fig:fold_auc_taxon_time}
\end{figure}

We compared the difference in AUC estimates from the out-of-sample forecasts to the AUC estimates from in-sample forecasts by subtracting the in-sample AUC estimates from the out-of-sample AUC estimates (Fig. \ref{fig:oos_ins_diff}); a difference in AUC close to 0 indicates complete congruence between the in-sample and out-of-sample forecasts. A positive difference indicates that out-of-sample forecasts actually outperform in-sample forecasts, whereas negative difference indicates poorer out-of-sample performance than in-sample forecast. Divergences between out-of-sample and in-sample forecasts are rare and tend not to cluster in time, consistent with the broad visual congruence between the in-sample and out-of-sample performance (Fig. \ref{fig:auc_taxon_time}, \ref{fig:fold_auc_taxon_time}). The only major multimillion year pattern indicating significantly poorer out-of-sample forecast performance than in-sample forecast performance is for Radiolaria based on the VP model concentrated around 30 Mya (Fig. \ref{fig:oos_ins_diff}).

\begin{figure}[ht]
 \centering
 \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/auc_diff}
 \caption{Comparing out-of-sample and in-sample forecasts. Congruence between in-sample and out-of-sample indicates that a model is not necessarily overfit to the data. This value is calculated as the values presented in Figure \ref{fig:fold_auc_taxon_time} minus those values presented in Figure \ref{fig:auc_taxon_time}. A differences close to 0 indicate complete congruence between in-sample and out-of-sample forecasts, while a positive difference indicates that the out-of-sample forecasts are actually higher performing than the in-sample forecasts, and a negative difference indicates poorer out-of-sample performance than in-sample forecast. See Table \ref{tab:model_def} for a description of each of the four models (C, V, CP, VP).}
 \label{fig:oos_ins_diff}
\end{figure}

@@

Three of the four models we evaluated are practically identical in their ability to make in-sample and out-of-sample forecasts. Although the in-sample AUC estimates differ between models, all of these estimates are in a narrow range of possible AUC values (Fig. \ref{fig:auc_hist}). The model with the best in-sample forecasting results includes the historical covariates and allows all covariate effects to vary over time. However, the out-of-sample forecasts from this model are biased, indicating that it is overfit to the data \ref{fig:fold_auc}). The model that includes historical covariates such as geographic range trajectory yields out-of-sample forecasts with nearly identical results to the model that allows covariate effects to vary over time but does not include historical covariates.

We note that both including historical covariates such as change in geographic range and allowing covariate effects to vary over time are different ways of encoding information from the past. Including historical covariates in a model but not allowing covariate effects to vary over time encodes the past explicitly but assumes that covariates effects are constant over time. Allowing covariate effects to vary over time, on the other hand, does not explicitly encode ``the past'' into our model but instead models how covariate effects vary over time which allows the past to be implicitly included in our model. By modeling this variation, forecasts made for future extinction events can allow for a wide range of potential effects of our covariates on predicting extinction. Comparing our out-of-sample forecast results indicates that these approaches yield approximately equal forecasting performance (Fig. \ref{fig:fold_auc}). Our results complement those of \citet{Kiessling2016} who found that including historical range trajectories improved the performance of extinction risk forecasts.

The relative quality and consistency between in-sample and out-of-sample forecasting performance for three of the four models is encouraging given that these estimates are based on very limited biological and environmental information about the studied taxa. Even our most complex models only account for a few simple aspects of geographic range, prior history, and phylogenetic affinity. The principal reason we were not able to include more biological information in the models used here is because we lack additional life history or ecological information for many of the marine micro- and nannoplankton included in this study. Foraminifera are an exception to this problem as aspects of life history, ecology, and physiology are known for many foram species \citep{Ezard2011}. However, comparable information does not exist for all foram species, nor does this type of data exist for the other three taxonomic groups studied here. Future analyses including this type of information and focused more narrowly on the Foraminifera may be informative.

An extremely important caveat, of course, is that human impacts may substantially alter present and future extinction risk dynamics relative to the average Cenozoic state, so that the future may become less predicatable than it has been in the past \citep{Harnik2012a,Finnegan2015}. The CP model assumes that the effects of the historical covariates are constant through time, but given growing evidence that human impacts substantially alter extinction risk dynamics \citep{Harnik2012a,Finnegan2015,Payne2007}, this assumption may not be valid and may limit or bias our ability to predict extinction in truly novel environmental regimes. Thus, it might be preferable to use a model similar to our V model which allows extinction risk and selectivity to vary over time. For this reason, while our CP and V models yield similar out-of-sample forecasts, we believe the V model offers more practical benefits for predicting extinction risk in future, anthropogenically impacted environments.

On a related note, it is notable that there are no obvious consistent changes in average model performance during episodes of climatic environmental change such as the mid-Eocene to early Oligocene (Fig. \ref{fig:fold_auc}), although this interval is characterized by elevated extinction among several planktonic groups \citep{Prothero1994,Wade2008,Kamikuri2012}. This suggests that relationships between the simple covariates included in our models and extinction risk were relatively stable though this interval.  However, it is likely that more complex models accounting for other aspects of ecology and biogeography might exhibit reduced forecasting performance during this and other climate change episodes.

To summarize, our results suggest that models trained on prior extinction/survival patterns do modestly well at predicting relative extinction probability of randomly selected species pairs based on a small number of simple taxonomic, geographic, and historical predictors. Although a model that includes historical covariates such as change in geographic range and change in climate between observations while also allowing covariate effects to vary over time performs best at in-sample prediction, this model is overfit to our data and produces less accurate out-of-sample forecasts than our three less complex models. The remaining three models yield nearly equivalent out-of-sample forecasts, suggesting that including historical information via either explicit modeling of historical covariate effects or modeling how covariate effects have changed over time does not diminish and may ultimately improve our ability to forecast future extinctions. The results of this simple exercise suggest that conservation decisions could indeed be bolstered by including fossil data.
